\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2013-03-14 doc/USER_MANUAL/manual_SPECFEM3D_Cartesian.tex
Added fault kinematics & dynamics as separate chapter to manual

>>>>

\chapter{\label{cha:Fault-Kinematics-Dynamics}Fault Kinematics \& Dynamics}

Fault is a discontinuity in the continuum that needs special treatment
with two nodes at the same locations. These are called split nodes
that allow to capture deformation on the fault plane termed slip.
Fault motion determines the deformation of the rest of the bulk. This
fault motion can be of two types either physics based spontaneous
rupture propagation or a prescribed rupture description. The former
is known as fault dynamics while the later is fault kinematics. We
describe here first the common features to both fault dynamics and
kinematics and then give detailed description of parameters required
for each of them separately.

Input files : Two additional input files are required in \texttt{DATA}
directory for faults with split nodes\\

\texttt{\textbf{Par\_file\_faults}}\texttt{ - }specialized parameters
required by fault modules (\emph{if this file doesn't exist, code
assumes that there are no faults})

\texttt{\textbf{FAULT\_STATIONS}}\texttt{ - }stations on the fault
where timeseries output of slip, sliprate and traction is required
\texttt{}~\\

Currently faults can only be run with \texttt{xdecompose\_mesh }and\texttt{
CUBIT. xmeshfem3D }isn't yet ready to handle faults.\\

Important module related to fault :

Decomposing mesh stage : $\mathtt{src/decompose\_mesh\_SCOTCH/fault\_scotch.f90}$

Generating databases stage : $\mathtt{src/generate\_databases/fault\_generate\_databases.f90}$

Solver stage : $\mathtt{src/specfem3D/fault\_solver\_common.f90}$

{\small \qquad{}}$\mathtt{src/specfem3D/fault\_solver\_kinematic.f90}$

{\small \qquad{}}$\mathtt{src/specfem3D/fault\_solver\_dynamic.f90}$\\

The features and the format of inputs and outputs of fault routines
are subject to change.\\

\section{Mesh Generation with Split Nodes}

Faults need to be handled in a special way during mesh generation.
A fault surface must lie at the interface between elements (the mesh
must honor the fault surfaces). Moreover, a fault is made of two surfaces
in contact. Each of these two surfaces needs a separate set of nodes.
This approach is known as \textquotedbl{}split nodes\textquotedbl{}.
To facilitate the mesh generation with split nodes in Cubit, we need
to separate the two fault surfaces by a small distance, effectively
creating a tiny opening of the fault. Note that only the interior
of the fault must be opened, its edges must remain closed (except
the edge on the free surface). The fault is automatically closed later
by SPECFEM3D.\\

Here is an example Cubit script to generate a mesh with split nodes
for a buried vertical strike-slip fault:\\
reset \\
brick x 10 y 10 z 10 \\
webcut volume all with plane xplane \\
webcut volume all with plane yplane \\
webcut volume all with plane xplane offset 3 \\
webcut volume all with plane zplane offset 3 \\
webcut volume all with plane zplane offset -3 \\
imprint all \\
merge all \\
unmerge surf 160 \\
mesh vol all \\
set node constraint off \\
node in surf 168 move X 0 Y 0.01 Z 0 \\
node in surf 160 move X 0 Y -0.01 Z 0\\

\begin{figure}[htbp]
\noindent \begin{centering}
\includegraphics[scale=0.3]{\lyxdot \lyxdot /src/MAIN_TRUNK/SPECFEM3D_Cartesian/doc/USER_MANUAL/figures/faultmesh}
\\
\includegraphics[scale=0.3]{\lyxdot \lyxdot /src/MAIN_TRUNK/SPECFEM3D_Cartesian/doc/USER_MANUAL/figures/surf168}
\includegraphics[scale=0.3]{\lyxdot \lyxdot /src/MAIN_TRUNK/SPECFEM3D_Cartesian/doc/USER_MANUAL/figures/splitnodes} 
\par\end{centering}

\caption{Screenshots of the CUBIT example for embedded fault with split nodes:
entire mesh decomposed into several volumes bounding the fault (top),
fault surface 168 is selected (bottom-left) and split nodes of the
fault illustrating only interior nodes of fault are split while those
on the edges of fault surface are touching each other (bottom-right).}

\label{fig:examples.splitnodes} 
\end{figure}

The Cubit scripts ({*}.jou and {*}.py) in the directory $\mathtt{EXAMPLES}$
generate more complicated meshes. The {*}.py files are Python scripts
that execute Cubit commands and use the Cubit-python interface for
SPECFEM3D (see next section). The Python language allows to define
and manipulate variables to parameterize the mesh. Alternatively,
the Python script can call a Cubit journal file ({*}.jou), which looks
like the example above. Variables can be defined and manipulated there
using the $\mathtt{APREPRO}$ language built in Cubit.\\

\texttt{Note}: you should avoid gaps in the list of indices of mesh
objects with the following Cubit command:

compress ids hex face edge node')

(otherwise you will get a segmentation fault during domain decomposition)\\

\section{Cubit-Python Scripts for Faults}

The mesh generated in Cubit needs to be processed and exported in
a format compatible with SPECFEM3D. This is achieved in the Python
scripts by calling the Python-Cubit interface functions defined in
the CUBIT directory: 
\begin{enumerate}
\item Function \textquotedbl{}$\mathtt{define\_bc}$\textquotedbl{} (or
similar ones) must be called to set up the absorbing boundaries database 
\item Function \textquotedbl{}$\mathtt{fault\_input}$\textquotedbl{} must
be called once for each fault to set up the fault database 
\item Function \textquotedbl{}$\mathtt{cubit2specfem3d.export2SESAME}$\textquotedbl{}
must be called at the very end of the script to export the mesh in
a SPECFEM3D format. 
\end{enumerate}
The functions in \#1 and \#3 are for bulk and are not documented here.
We focus here on \#2:\\

\textbf{Function}: $\mathtt{fault\_input}$

\textbf{Purpose}: export a fault mesh database from Cubit to a SPECFEM3D-compliant
file

\textbf{Syntax}: fault\_input(id\_fault, ids\_surf\_1, ids\_surf\_2)\\

\textbf{Inputs}: id\_fault integer index assigned to the fault.

(The user must number all the faults, starting at 1, with unit increments) 
\begin{lyxlist}{00.00.0000}
\item [{ids\_surf\_1}] list of Cubit ids of all surfaces that form side
1 of the fault
\item [{ids\_surf\_2}] list of Cubit ids of all surfaces that form side
2 of the fault 
\end{lyxlist}
(The user must decide which side of the fault is side 1. This choice
affects the sign conventions of fault quantities as explained in a
later section)\\

\textbf{Outputs}: file \textquotedbl{}fault\_file\_X.dat\textquotedbl{},
where X is the fault id (id\_fault).\\

\textbf{Example}: For the example in the previous section:\\

$\hphantom{}$A1 = {[}168{]}

$\hphantom{}$A2 = {[}160{]}

$\hphantom{}$faultA = fault\_input(1,A1,A2)\\

\section{Running a Simulation }
\begin{enumerate}
\item Copy the input files to the work directory:

>\textcompwordmark{}> cd \textasciitilde{}/$\mathtt{SPECFEM3D}$$\mathtt{}$

>\textcompwordmark{}> cp -r $\mathtt{EXAMPLES/splay\_faults/DATA}$
./

\item Create a mesh with CUBIT: (If you have not installed CUBIT yet skip
this step and use the mesh files included with the examples.)

See the previous section \textquotedbl{}Mesh Generation with Split
Nodes\textquotedbl{} or move to the directory CUBIT, open the cubit
GUI, run a CUBIT script: >\textcompwordmark{}> cd \textasciitilde{}$\mathtt{/SPECFEM3D/CUBIT}$
>\textcompwordmark{}> cubit In CUBIT's menu \textquotedbl{}Tools\textquotedbl{},
select \textquotedbl{}Play Journal File\textquotedbl{} and select
a script file, for instance: $\mathtt{EXAMPLES/splay\_faults/splay\_faults.py}$

This creates several mesh files in directory $\mathtt{CUBIT/MESH/}$:
absorbing\_surface\_file\_{*} (5 files) free\_surface\_file materials\_file
mesh\_file nodes\_coords\_file nummaterial\_velocity\_file

The CUBIT graphics window should show the mesh (e.g. $\mathtt{EXAMPLES/splay_{f}aults/splay_{f}aults.jpg}$)
You can zoom, pan or rotate using mouse gestures. To look at the block
of your interest, set \textquotedbl{}visibility off\textquotedbl{}
in the other blocks.

To run CUBIT from different working directories it is convenient to
include the path to $\mathtt{SPECFEM3D/CUBIT}$ in an environment
variable $\mathtt{PYTHONPATH}$

\item Partition the mesh with the domain decomposition software SCOTCH.

>\textcompwordmark{}> cd \textasciitilde{}/$\mathtt{SPECFEM3D/decompose\_mesh\_SCOTH}$

>\textcompwordmark{}> ./$\mathtt{decompose\_mesh\_SCOTCH}$ 'nproc'
../$\mathtt{CUBIT/MESH/}$ ../$\mathtt{DATABASES\_MPI}$ /

where 'nproc' is the number of processors that you will use in the
simulation. The second argument (../$\mathtt{CUBIT/MESH/}$ in this
example) is the path to the directory containing the mesh files that
were generated by CUBIT. This creates one mesh file proc000{*}{*}{*}\_Database
per processor in directory $\mathtt{DATABASES\_MPI}$ . These files
can be large, so you might want to place $\mathtt{DATABASES\_MPI}$
in a scratch disk and specify the path accordingly when executing
$x\mathtt{decompose\_mesh\_SCOTCH}$.

\item Edit the file $\mathtt{DATA/Par_{f}ile}$: (copy an example here,
see III.2)

$\mathtt{LOCAL\_PATH}$= should be the path to $\mathtt{DATABASES\_MPI}$

$\mathtt{NPROC}$= number of processors. The same as the number of
partitions in SCOTCH (step 3).

\item Generate databases:

>\textcompwordmark{}> cd \textasciitilde{}/$\mathtt{SPECFEM3D}$

>\textcompwordmark{}> mpirun -np nproc ./$\mathtt{xgenerate_{d}atabases}$

or

>\textcompwordmark{}> $\mathtt{./run/run.xdatabases}$

or

>\textcompwordmark{}> qsub -l nodes=... -l walltime=... go\_mesher
(the script go\_mesher is in $\mathtt{utils/}$)

This creates several binary mesh files for each processor (proc000{*}{*}{*}.bin)
in directory $\mathtt{DATABASES\_MPI}$.

\item If this is the first time you run a simulation with this mesh, set
the parameter DT (the time step) DATA/Par\_file to a value smaller
than the \textquotedbl{}Maximum suggested time step\textquotedbl{}
in $\mathtt{OUTPUT\_FILES/output\_mesher.txt}$. Consider the note
in section IV about the effect of Kelvin-Voigt damping on the critical
timestep. Edit the variables $\mathtt{NSTEP}$ and $\mathtt{NTSTEP\_*}$
in $\mathtt{DATA/Par\_file}$ accordingly.
\item Run the solver:

>\textcompwordmark{}> mpirun -np nproc ./$\mathtt{xspecfem3D}$

or

>\textcompwordmark{}> $\mathtt{./run/run.xspecfem3d}$

or

>\textcompwordmark{}> qsub -l nodes=... -l walltime=... go\_solver
(the script go\_solver is in $\mathtt{utils/}$)

\end{enumerate}

\section{Examples}

The package includes examples, the SCEC benchmark problems: 
\begin{itemize}
\item TPV5, a planar vertical strike-slip fault 
\item TPV14 and TPV15, vertical strike-slip fault system with a fault branch 
\item TPV102 and TPV103, rate-and-state benchmarks
\item TPV16, heterogenous initial stress
\item Splay fault models from Wendt et al. (2009)\\

\end{itemize}
To run the examples:
\begin{enumerate}
\item Read the documents in the directory $\mathtt{EXAMPLES/*/description}$.
They contain a description of the example and additional instructions
to run it. 
\item Replace the contents of directory $\mathtt{DATA/*}$ by one of the
$\mathtt{EXAMPLES/tpv*/DATA}$, for instance:

>\textcompwordmark{}> cp -r $\mathtt{EXAMPLES/tpv5/DATA}$ ./

\item Follow all the steps in section II above.
\item Visualize the results with the matlab scripts in the directory $\mathtt{EXAMPLES/*/post}$
\end{enumerate}

\section{Sign Convention for Fault Quantities}

During mesh generation, the fault is defined by two surfaces in contact.
Let's denote as \textquotedbl{}\emph{side 1}\textquotedbl{} the SECOND
surface declared by the user in the call to the python function \textquotedbl{}\emph{fault\_input}\textquotedbl{},
and the FIRST surface as \textquotedbl{}\emph{side 2}\textquotedbl{}.
The local coordinate system on the fault is defined as the right-handed
coordinate system defined by (strike, dip, normal), where \textquotedbl{}normal\textquotedbl{}
is the normal vector outgoing from side 1, \textquotedbl{}\emph{dip}\textquotedbl{}
is parallel to the along-dip direction pointing downwards, and \textquotedbl{}\emph{strike}\textquotedbl{}
is the horizontal along-strike vector such that the system is right-handed.\\

Slip is defined as displacement on side 2 minus displacement on side
1. In the local coordinate system on the fault, positive along-strike
slip is right-lateral and positive along-dip slip is thrust if side
1 is on the hanging wall (normal faulting if side 1 is on the foot
wall).\\

Traction is defined as the stress induced on side 1 by side 2, which
is the stress tensor times the normal vector outgoing from side 1.
In the local coordinate system on the fault, the normal traction is
negative in compression, positive along-strike traction generates
right-lateral slip and positive along-dip traction generates thrust
slip if side 1 is on the hanging wall (normal faulting if side 1 is
on the foot wall).

\section{Input Files}

\textbf{DATA/Par\_file} 
\begin{lyxlist}{00.00.0000}
\item [{\textbf{DATA/STATIONS}}] List of stations outside the fault 
\item [{\textbf{DATA/Par\_file\_faults}}] contains parameters of the fault.
The first part of this file has a strict format:

\begin{description}
\item [{Line}] 1: Number of faults (NF) 
\item [{Lines}] 2 to NF+1: Kelvin Voigt damping (in seconds) for each fault.
(See below how to set this parameter) 
\item [{Line}] NF+2: Type of simulation (1=dynamic , 2 = kinematic) 
\item [{Line}] NF+3: Number of time steps between updates of the time series
outputs at selected fault points (see DATA/FAULT\_STATIONS), usually
a large number (100s or 1000s). Note that the sampling rate of the
time series is usually much higher.
\item [{Line}] NF+4: Number of time steps between fault snapshot outputs
(quantities at every fault point exported at regular times), usually
a large number (100s or 1000s). 
\item [{Line}] NF+5: Slip velocity threshold below which frictional healing
is set (friction coefficient is reset to its static value). If this
value is negative healing is disabled. 
\item [{Line}] NF+6: Slip velocity threshold to define the rupture front.
Only used for outputs.
\end{description}

The rest of this file is made of namelist input blocks (see \textquotedbl{}namelist\textquotedbl{}
in a Fortran 9x manual). The input for each fault has the following
sequence (arguments in {[}brackets{]} are optional):

\end{lyxlist}
\&\textbf{BEGIN\_FAULT} /

\&\textbf{INIT\_STRESS} S1, S2, S3 {[},n1, n2, n3{]} /

followed by (n1+n2+n3) \&\textbf{DIST2D} blocks

\&\textbf{SWF} mus, mud, dc {[}, nmus, nmud, ndc{]} /

followed by (nmus+nmud+ndc) \&\textbf{DIST2D} blocks\\

\begin{lyxlist}{00.00.0000}
\item [{\&\textbf{INIT\_STRESS}}] input block sets the initial fault stresses
relative to the foot-wall side of the fault. Initial stresses are
composed of a constant background value possibly overwritten in prescribed
regions by heterogeneous distributions (see \&\textbf{DIST2D} blocks
below):

S1 = constant background value of along-strike shear stress (positive
in the usual strike direction)

S2 = constant background value of along-dip shear (positive is down-dip,
normal faulting)

S3 = constant background value of normal stress (negative in compresion)

n1 = number of heterogeneous items for along-strike shear stress {[}default
is 0{]}

n2 = number of heterogeneous items for along-dip shear stress {[}default
is 0{]}

n3 = number of heterogeneous items for normal stress {[}default is
0{]}\\

\item [{\&\textbf{SWF}}] input block sets the slip-weakening friction parameters
of the fault:

mus = constant background value of static friction coefficient

mud = constant background value of dynamic friction coefficient

dc = constant background value of critical slip-weakening distance

nmus = number of heterogeneous items for static friction coefficient
{[}default is 0{]}

nmud = number of heterogeneous items for dynamic friction coefficient
{[}default is 0{]}

ndc = number of heterogeneous items for critical slip-weakening distance
{[}default is 0{]}\\

\item [{\&\textbf{DIST2D}}] input blocks modify (overwrite) the value of
a fault parameter by a heterogeneous spatial distribution:

\&\textbf{DIST2D} shape='$\mathtt{square}$', val, xc, yc, zc, l /

$\;$sets a constant value (val) within a cube with center (xc,yc,zc)
and edge size l.

\&\textbf{DIST2D} shape='$\mathtt{rectangle}$', val, xc, yc, zc,
lx, ly, lz /

$\;$sets a constant value (val) within a rectangular prism with center
(xc,yc,zc) and edge sizes (lx,ly,lz).

\&\textbf{DIST2D} shape='$\mathtt{rectangle-taper}$', val, valh,
xc, yc, zc, lx, ly, lz /

$\;$sets a vertical linear gradient within a rectangular prism with
center (xc,yc,zc) and edge sizes (lx,ly,lz). Values vary linearly
as a function of vertical position z between value val at z = zc-lz/2
and value valh at z = zc+lz/2 .

\&\textbf{DIST2D} shape='$\mathtt{circular}$', val, xc, yc, zc, r
/

$\;$sets a constant value (val) within a sphere with center (xc,yc,zc)
and radius r. \\

\item [{\textbf{DATA/FAULT\_STATIONS}}] Stations in the fault plane.

Line 1: number of stations.

Line 2 to end: 5 columns: X, Y, Z (-depth), station name, fault-id

The fault-id identifies the fault that contains the station. It is
the index of appearance in the faults list after line 2 of Par\_file\_faults

\item [{\textbf{DATA/input\_file.txt}}] Heterogeneous stresses and friction
parameters Documented in page 10 of EXAMPLES/tpv16.crack/description/TPV16\_17\_Description\_v03.pdf

To activate this feature, in fault\_solver.f90 set TPV16 = .true.
then re-compile the code: cd SPECFEM3D make

\item [{Heterogeneous}] velocity models can be given in a regular grid.

In the mesh generation Python file (.py) or in the CUBIT journal file
(.jou), set the material 1st attribute to -1:

\qquad{}block {*} attribute index 1 -1

In module model\_tomography.f90:

\qquad{}Set the parameter TOMO\_FILENAME to the name of the file
containing the velocity model.

Re-compile the code:

\qquad{}cd SPECFEM3D

\qquad{}make \\

The format of the velocity file is: 
\begin{description}
\item [{Line}] 1: ORIG\_X, ORIG\_Y, ORIG\_Z, END\_X, END\_Y, END\_Z (coordinates
of the two extreme corners of the box) 
\item [{Line}] 2: SPACING\_X, SPACING\_Y, SPACING\_Z (regular grid spacing
in each direction) 
\item [{Line}] 3: NX, NY, NZ (number of grid point in each direction) 
\item [{Line}] 4: VP\_MIN, VP\_MAX, VS\_MIN, VS\_MAX, RHO\_MIN, RHO\_MAX
(min and max values of P wave speed, S wave speed and density) 
\item [{Line}] 5 to 4+NX{*}NY{*}NZ: x\_tomo,y\_tomo,z\_tomo,vp\_tomo,vs\_tomo,rho\_tomo
(position, P wave speed, S wave speed, density)\\

\end{description}
\end{lyxlist}
Several files are generated automatically by xgenerate\_databases
in directory DATABASES\_MPI and do not need to be modified by the
user.

\section{How to set the Kelvin-Voigt Damping Parameter }

The purpose of the Kelvin-Voigt viscosity in the dynamic fault solver
is to damp spurious oscillations generated by the fault slip at frequencies
that are too high to be resolved by the mesh. The viscosity \textquotedbl{}\emph{eta}\textquotedbl{}
(in seconds) depends on the size of the elements on the fault. Here
is how to set it:
\begin{enumerate}
\item Determine the average linear size of the elements on the fault plane,
\textquotedbl{}\emph{h\_fault}\textquotedbl{}. Usually this value
is prescribed by the user during mesh generation. Otherwise it can
be found by inspection of the mesh inside the Cubit GUI. 
\item Use the matlab function $\mathtt{utils/critical_{t}imestep.m}$ to
compute dtc\_fault = critical\_timestep(cp,h\_fault,ngll) This is
the critical time step in an elastic medium for a hypothetical element
of cubic shape with size equal to h\_fault. 
\item Set eta in Par\_file\_faults to (0.1 to 0.3){*}dtc\_fault. A larger
eta damps high-frequencies more aggresively but it might also affect
lower frequencies and rupture speed.

Viscosity reduces numerical stability: the critical timestep in a
simulation with Kelvin-Voigt damping needs to be smaller than that
in a purely elastic simulation. Here is how to set the time step accordingly:

\item Run a test simulation without viscosity (eta=0 and only a few time
steps) 
\item Look for the \textquotedbl{}maximum suggested time step\textquotedbl{}
in $\mathtt{OUTPUT_{F}ILES/output_{m}esher.txt}$ This is the critical
timestep of a purely elastic simulation, \textquotedbl{}dtc\_bulk\textquotedbl{}. 
\item Reset the timestep of the simulation with a Kelvin-Voigt material
to a value smaller than dtc\_kv = eta{*}( sqrt(1+dtc\_bulk\textasciicircum{}2/eta\textasciicircum{}2)-1
)
\end{enumerate}
Note that in general dtc\_bulk is smaller than dtc\_fault, because
elements off the fault might be smaller or more distorted than element
faces on the fault.

\section{Output Files }

Several output files are saved in \textasciitilde{}/$\mathtt{SPECFEM3D/OUTPUT_{F}ILES}$:\\

\begin{enumerate}
\item Seismograms for each station on the fault plane given in $\mathtt{DATA/FAULT_{S}TATIONS}$.
One output file is generated for each station, named after the station.
The files are ascii and start with a header (22 lines long) followed
by a data block with the following format, one line per time sample:\\

\# Column \#1 = Time (s)

\# Column \#2 = horizontal right-lateral slip (m)

\# Column \#3 = horizontal right-lateral slip rate (m/s)

\# Column \#4 = horizontal right-lateral shear stress (MPa)

\# Column \#5 = vertical up-dip slip (m)

\# Column \#6 = vertical up-dip slip rate (m/s)

\# Column \#7 = vertical up-dip shear stress (MPa)

\# Column \#8 = normal stress (MPa)\\

The stresses are relative to the footwall side of the fault (this
convention controls their sign, but not their amplitude). Slip is
defined as displacement of the hanging wall relative to the footwall.

\item Seismograms at stations in the bulk (out of the fault plane) given
in $\mathtt{DATA/STATIONS}$. 
\item Rupture time files are named Rupture\_time\_FAULT-id. One file is
generated for each fault. The files are ascii and start with a header
(12 lines long) followed by a data block with the following format,
one line per fault node:

\# Column \#1 = horizontal coordinate, distance along strike (m)

\# Column \#2 = vertical coordinate, distance down-dip (m)

\# Column \#3 = rupture time (s)

\item Fault quantities (slip, slip rate, stresses, etc) at regular times
are stored in binary data files called Snapshot\#it\#.bin, where \#it\#
is the timestep number. These can be read in Matlab with the function
$\mathtt{Post-processing/FSEM3D\_snapshot.m}$
\end{enumerate}

\section{Post-processing and Visualization }

Some Matlab functions for post-processing and visualization are included
in directory Post-processing. The functions are internally documented
(see their matlab help).\\

\texttt{FSEM3D\_snapshot} reads a fault data snapshot\\

The directories$\mathtt{EXAMPLES/*/post}$ contain additional Matlab
scripts to generate figures specific to each example.
